\section{Semantics of First-Order Logic}

We present the semantics using basis calculus.

\subsection{Worlds}

Worlds in First-Order Logics contain varying numbers of objects.
We use domain variables $\indvariable$ to enumerate the sets of objects.

Worlds are represented by three sets of tensors:
\begin{itemize}
    \item One-hot encodings of constants
    \item Coordinate encodings for each predicate
    \item Basis encoding of each function
\end{itemize}

\subsection{Basis encodings of terms}

Terms are object valued functions, which $r_{\folterm}$ different arguments are also object valued and called variables.
We represent terms by the corresponding basis encodings
\begin{align*}
    \bencodingofat{\folterm}{
        \indvariableof{\folterm}, \indvariableof{[r_{\folterm}]}
    }
\end{align*}

Basic blocks:
\begin{itemize}
    \item Constant
    \stantikzpic{
        \draw[->-] (0,1) -- (0,2.5) node[midway,right] {\colorlabelsize $\indvariableof{\folterm}$};
        \drawsquaretensor{0}{0}{$\onehotmapof{\indindex}$}
    }
    \item Variable
    \stantikzpic{
        \draw[->-] (0,1) -- (0,2.5) node[midway,right] {\colorlabelsize $\indvariableof{\folterm}$};
        \drawsquaretensor{0}{0}{$\identity$}
        \draw[->-] (0,-2.5) -- (0,-1) node[midway,right] {\colorlabelsize $\indvariable$};
    }
    \item Function of terms
    \stantikzpic{
        \draw[->-] (2,1) -- (2,2.5) node[midway,right] {\colorlabelsize $\indvariableof{\folterm}$};
        \draw (-1,-1) rectangle (5,1);
        \drawtextnode{2}{0}{\corelabelsize $\bencodingof{\exfunction}$}
        \draw[->-] (0,-2.5) -- (0,-1) node[midway,right] {\colorlabelsize $\indvariableof{0}$};
        \draw[->-] (4,-2.5) -- (4,-1) node[midway,right] {\colorlabelsize $\indvariableof{r_{\exfunction}}$};
    }
\end{itemize}
Contractions of those terms along directed acyclic graphs with single root nodes represent more complex terms.
As such, basis encodings of functions can be contracted with constants, variables or other functions to build up basis encodings of more complex terms.

\subsection{Atomic sentences}

Atomic sentences are contractions of coordinate encodings of predicates with basis encodings of terms.

\subsection{Complex sentences}

Quantifier free sentences are contractions of basis encodings to atomic sentences with basis encodings of the connectives.

Quantifiers act as non-linear transforms of tensors.