\documentclass[aps,onecolumn,nofootinbib,pra]{article}

\usepackage{../../arxiv}
\usepackage{amsmath,amsfonts,amssymb,amsthm,bbm,graphicx,enumerate,times}
\usepackage{mathtools}
\usepackage[usenames,dvipsnames]{color}
\usepackage{hyperref}
\hypersetup{
	breaklinks,
	colorlinks,
	linkcolor=gray,
	citecolor=gray,
	urlcolor=gray,
	pdftitle={The Tensor Network Approach to Efficient and Explainable AI},
	pdfauthor={Alex Goessmann}
}

\usepackage{tikz}
\usepackage{graphicx}
\usepackage{float}
\usepackage{comment}
\usepackage{csquotes}

\usepackage{listings}
\usepackage{verbatim}
\usepackage{etoolbox}
\usepackage{braket}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{titlesec}
\usepackage{tikz}
\usepackage{mathtools}
\usepackage{fancyhdr}
\usepackage{bbm}
\usepackage{bm}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{lipsum}

\newtheorem{remark}{Remark}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\input{../../macros/general_macros.tex}
\input{../../macros/tc_macros.tex}
\input{../../macros/tikz_macros.tex}

\pretolerance=500
\tolerance=100 
\emergencystretch=10pt

% Bibliography
\DeclareUnicodeCharacter{FB01}{fi}
\usepackage[round]{natbib}


\newcommand{\red}[1]{\textcolor{red}{#1}}

\begin{document}
\title{\tnreason{} for LLM Reasoning}

\maketitle
\date{\today}

We here sketch, how the usage of \tnreason{} can be enhanced based on LLMs, and how \tnreason{} could be exploited in the retraining of LLMs.

\section{Copilots for the usage of \tnreason{}}

We envision users to provide prompts describing their reasoning objectives, which the LLM interprets to generate precise tnreason function calls, ensuring a natural and efficient workflow.
By leveraging LLMs' natural language understanding, users can describe complex reasoning tasks intuitively, allowing the copilot to translate them into structured calls to \tnreason{}.

\subsection{Model explainability}

% Explainability
Having a \tnreason{} model described in the formal script language, a LLM can be prompted to explain that model to an user.
Besides datapoint specific inquiries, such a copilot can be used to explain entire parts of the model, for example based on a collection of rules triggered when having a specific input feature.
The envisioned explainability therefore goes beyond the typical "post-hoc" explainability towards an intrinsic "ante-hoc" explainability bringing guarantees for a models behavior.

\subsection{Prompt structure}

% Prompt structure
The systems prompt will describe the basic functionality of \tnreason{} and the users prompt consists in the requirements of the use case, formulated in natural language.
Here the user does not have to be informed about the formal grammar of the \tnreason{} script language, nor the structure and expressivity of propositional logics.
This approach enhances accessibility of \tnreason{}, enabling seamless interaction between human intuition and formal logical computation.

% Gemini
Preliminary experiments with the Gemini model on Colab can be found here:
\url{https://colab.research.google.com/drive/1ESzJeZp7O022I1z31NeA8VExALUVfUf7?usp=sharing}

\subsection{Multi-agent Orchestration}

Another interesting idea is a multi-agent orchestration, where the ability of LLMs and formal proof languages are balanced with respect to each other.
Here one might exploit the LLMs as a translator between human understandable text and formal statements such as in the \tnreason{} script language.
\begin{itemize}
	\item \textbf{Communication agent (LLM):}
		Interacts with the user, to create and maintain a \tnreason{} model.
		Further, when asked about the models behavior at a specific data instance or in a more generic situation, it interprets the model in human understandable language.
		At best realized by a LLM to handle the flexible user-specific prompts.
	\item \textbf{Execution agent (\tnreason{}):}
		Maintains a model about a use case in cooperation with the communication agent, and executes the inference calls from the data agent.
		At best realized in an efficient and explainable manner and therefore in a formal, declarative language such as the \tnreason{} script language.
	\item \textbf{Data agent (Classical Software):}
		Handles the runtime of a productive model, by transforming the input data into an inference instance, passing it to the execution agent, and preparing the output in a specific format.
\end{itemize}

\section{Retraining LLMs based on a reinforcement environment created by \tnreason{}}

Retraining LLMs in a reinforcement learning environment incorporating \tnreason{} workloads enables adaptive improvement in logical reasoning and decision-making.
The model receives feedback based on the correctness and efficiency of its generated \tnreason{} calls, refining its ability to solve complex reasoning tasks over time.
This iterative training approach enhances the LLM’s alignment with formal reasoning principles, leading to more reliable and interpretable outputs.

% Reward
One can pose \tnreason{} inference models, such as probabilistic queries of formulas in a hybrid knowledge base generated from random, as a prompt and calculate the reward as the absolute difference of the exact calculation by \tnreason{} and the response from the LLM.


\section{Ideas to Monte Carlo Tree Searches}

Monte Carlo Tree Search (MCTS) is a classical method in AI to solve adversarial games.
Recent work \cite{besta_reasoning_2025,xin_deepseek-prover-v15_2024} integrate them into the architecture of LLMs to enhance reasoning capabilities.


\subsection{Parametrizing a Search Tree by a Formula Selecting Network}

Search Trees in combination with Monte Carlo tree searches MCTS received recent popularity as a mechanism to advanced reasoning functionalities of LLMs.

In the \tnreason{} framework, several formula selecting networks (and other selection architectures such as slice selection) follow a similar idea.
Instead of branches, we typically depict choices by assignments to selection variables.
Thus, a formula selecting network parametrizes formulas, which are in the expressivity of formula search trees, where the branches along each layer of the tree have a common structure (such as selection of a logical connective or selection of a categorical variable).



\subsection{Solving mode search problems by MCTS}

Mode search problems have been a central class of problems in the main documentation, ranging from MAP inference of \MarkovLogicNetworks{} to the grafting heuristic in greedy learning.
So far, we have investigated methods from variational inference as well as reformulations into known optimization formats (HUBO) to solve mode search problems.
MCTS provide alternative ways to solve such search problems, which follow a sampling based approach and make use of the search tree parametrization in the section above.
Let us note, that while MCTS could operate on any tree representing a search space, we here have specific trees with constant branching factors to the nodes in each layer, and with each leaf corresponding with a coordinate.

The decisions, which have to be done iteratively, are assignments to the variables $\shortcatvariables$ on the axis of a tensor $\energytensor$.
Given assignments $\catindexof{[\catenumerator]}$ to the first $\catenumerator$ variables $\catvariableof{[\catenumerator]}$, we could choose an assignment on $\catvariableof{\catenumerator}$ based on 
\begin{align*}
	\argmax_{\catindexof{\catenumerator}\in[\catdimof{\catenumerator}]} \quad \contractionof{\energytensorat{\shortcatvariables}}{\indexedcatvariableof{[\catenumerator]},\indexedcatvariableof{\catenumerator}} \, .
 \end{align*}
 For efficiency reasons, MCTS optimizes an approximative problem instead, where the above contraction is estimated by a particle-based approximation of the tensor.

Alternatives here are further sampling approaches considered in \tnreason{}, such as Gibbs sampling strategies resampling previous parameters in combination with simulated annealing.

When annealing the underlying distribution by scaling the energy tensor, the search objective in each decision step is more sensitive to the maximum in the respective slice, but harder to estimate (since the maximum is typically not found during estimation).

\section{Literature}

Small LLMs suffice to develop reasoning capability:
\begin{itemize}
	\item \cite{guan_rstar-math_2025}
\end{itemize}

Reinforcement learning to enhance the reasoning capability of LLMs:
\begin{itemize}
	\item \cite{deepseek-ai_deepseek-r1_2025}
\end{itemize}

Combination with the LEAN language:
\begin{itemize}
	\item \cite{xin_deepseek-prover-v15_2024}
\end{itemize}

Monte Carlo Tree Search:
\begin{itemize}
	\item \cite{besta_reasoning_2025}: Integration of MCTS into Reasoning Language Models (RLM)
	\item \cite{xin_deepseek-prover-v15_2024}: RMaxTS algorithm
\end{itemize}


\bibliographystyle{plainnat}
\bibliography{llm_reasoning}

\end{document}