\documentclass{beamer}

\input{../../spec_files/course_preamble.tex}
\input{../../spec_files/course_macros.tex}

\title[Propositional Formulas]{
    \coursechapterone \\
    {\huge Propositional Formulas}
}

\begin{document}

{\frame[plain]{\titlepage}}
    \showoutlineatsection


    \section{Propositional Logics}

    \begin{frame}{Syntax and Semantics of Logics}

        Logic is build upon
        \textbf{Syntax}
        \begin{itemize}
            \item How to formulate logical formulas?
            \item Example: The formula $\var{Feature}\Rightarrow\var{Account 1}$ represents the implication, that the $\var{Account 1}$ has to be booked when the feature $\var{Feature}$ is present on the invoice.
            \item Propositional Logics: The variables (or atomic formulas) are combined with logical connectives $\{\lnot, \land, \lor, \Rightarrow, \Leftrightarrow\}$ to build generic formulas.
        \end{itemize}
        and \textbf{Semantics}
        \begin{itemize}
            \item What do these formulas describe?
            \item Example: From the formula $\var{Feature}\land(\var{Feature}\Rightarrow\var{Account 1})$ the formula $\var{Account 1}$ follows.
            \item Propositional Logics: Each formula describes possible worlds, which are called models of a formula. Comparing the models defines the \textbf{entailment relation}.
        \end{itemize}

    \end{frame}

    \begin{frame}{States of an atomic variable}

        Any atomic variable $\catvariable$ (e.g. the feature $\catvariableof{F}$ present on the invoice) can be in two states
        \begin{align*}
            \mathrm{False} \Leftrightarrow 0 \quad,\quad \mathrm{True} \Leftrightarrow 1
        \end{align*}
        We understand the states as the coordinates of a vector:
        \begin{center}
            \begin{tikzpicture}[scale=1]
                \node (A) at (0,0) {
                    $\coloredmatrixof{\cdot \\ \cdot }{\catvariable}$
                };
                \draw[dashed,->] (-0.5,0.2) to (-1.5,0.2) node[left]{\small $\mathrm{False}$};
                \draw[dashed,->] (-0.5,-0.2) to (-1.5,-0.2) node[left]{\small $\mathrm{True}$};
            \end{tikzpicture}
        \end{center}

        \textbf{Conventions:}
        \begin{itemize}
%            \item We assign to each state a coordinate of a vector, in which we can store whether the variable can be in the state
            \item We assign a $0$ in the coordinate, when the corresponding state is not possible, and a $1$ if the corresponding state is possible.
            \item In the brackets after a vector we denote the corresponding variable.
        \end{itemize}


    \end{frame}

    \begin{frame}{Encoding knowledge about a variable}

        The knowledge about an atomic variable is aligned in an two-dimensional vector:
        \begin{itemize}
            \item When the variable is known to be $\mathrm{True}$:
            \begin{align*}
                \coloredmatrixof{0\\1}{\catvariable}
            \end{align*}
            \item When the variable is known to be $\mathrm{False}$:
            \begin{align*}
                \coloredmatrixof{1\\0}{\catvariable}
            \end{align*}
            \item When the state of the variable is unknown:
            \begin{align*}
                \coloredmatrixof{1\\1}{\catvariable}
            \end{align*}
            \item The variable is neither $\mathrm{True}$ nor $\mathrm{False}$ (our knowledge is then inconsistent)
            \begin{align*}
                \coloredmatrixof{0\\0}{\catvariable}
            \end{align*}
        \end{itemize}

    \end{frame}


    \begin{frame}{Knowledge about two variables}
        Let us extend the system to two variables:
        \begin{itemize}
            \item $\catvariableof{A1}$: Whether the $\var{Account 1}$ is booked.
            \item $\catvariableof{A2}$: Whether the $\var{Account 2}$ is booked.
        \end{itemize}
        The states of the tuple $(\catvariableof{A1},\catvariableof{A2})$ are the product
        \begin{align*}
            \{0,1\} \times \{0,1\} = \{(0,0),(0,1),(1,0),(1,1)\}
        \end{align*}
        For example $(\catvariableof{A1},\catvariableof{A2})=(1,0)$ refers to the system state, in which $\var{Account 1}$ is booked ($\mathrm{True}$) and $\var{Account 2}$ is not booked ($\mathrm{False}$).

        \begin{center}
            \begin{tikzpicture}[scale=1]
                \node (A) at (0.7,0) {
                    $\coloredmatrixof{\cdot & \cdot \\ \cdot & \cdot}{\catvariableof{A1},\catvariableof{A2}}$
                };
                \draw[<-,dashed] (-0.9,-0.275) node[right] {\tiny $1$} -- (-0.9,0.275) node [midway, left] {\tiny $\catvariableof{A1}$} node[right] {\tiny $0$};
                \draw[->,dashed] (-0.3,0.85) node[below] {\tiny $0$} -- (0.3,0.85) node [midway, above] {\tiny $\catvariableof{A2}$} node[below] {\tiny $1$};
            \end{tikzpicture}
        \end{center}

    \end{frame}

    \begin{frame}{Knowledge about two variables}
        The rule, that exactly one of the accounts $\var{Account 1}$ and $\var{Account 2}$ is booked is then encoded in the matrix
        \begin{center}
            \begin{tikzpicture}[scale=1]
                \node (A) at (0,0) {
                    $\coloredmatrixof{0 & 1 \\ 1 & 0}{\catvariableof{A1},\catvariableof{A2}}$
                };
                \draw[dashed,->] (-1.5,0.2) to (-2.5,0.2) node[left]{\small $(\mathrm{False},\mathrm{False})$};
                \draw[dashed,->] (-1.5,-0.2) to (-2.5,-0.2) node[left]{\small $(\mathrm{True},\mathrm{False})$};

                \draw[dashed,->] (-0.5,0.5) to (-0.5,1) node[above]{\small $(\mathrm{False},\mathrm{True})$};
                \draw[dashed,->] (-0.5,-.5) to (-0.5,-1) node[below]{\small $(\mathrm{True},\mathrm{True})$};

            \end{tikzpicture}
        \end{center}

        The rule that $\var{Account 1}$ is booked when $\var{Feature}$ is present by:
        \begin{align*}
            \coloredmatrixof{1 & 1 \\ 0 & 1}{\catvariableof{F},\catvariableof{A1}}
        \end{align*}

    \end{frame}


    \begin{frame}{Knowledge about the Toy Accounting System}
        Now let us extend to the toy accounting system.
        Given the Knowledge about both formulas we have
        \begin{center}
            \input{tikz-pics/toy-accounting-hard}
        \end{center}

        To understand the transition from the matrices to this tensor of order $3$, we need to widen our tensor formalism.

    \end{frame}


    \section{Coarser Picture: Tensors}

    \begin{frame}{Situation of $\atomorder$ variables}
        Let us now imagine that we have $\atomorder$ variables $\catvariableof{0},\ldots,\catvariableof{\atomorder-1}$.

        The coordinatewise description would be confusing for $\atomorder>3$!

        The states are the set
        \begin{align*}
            \bigtimes_{\catenumeratorin} \{0,1\}
        \end{align*}
        of $2^{\catorder}$ state tuples.

        We use a diagramatic depiction instead:


    \end{frame}

    \begin{frame}{Propositional Formulas}
        \begin{definition}[Propositional Formula]
            A \emph{propositional formula} of order $\atomorder\in\nn$ is a map
            \begin{align*}
                \exformula : \bigtimes_{\catenumeratorin} \{0,1\} \rightarrow \{0,1\} \, .
            \end{align*}
%            \[ \hypercore : \facstates \rightarrow \rr \, . \]
            \begin{itemize}
                \item We refer to $\hypercoreat{\indexedshortcatvariables}\in\rr$ by the \emph{coordinate} of $\hypercore$ with the indices $\shortcatindices=\catindices$.
                \item By scalar multiplication and addition of maps, tensors build linear spaces called \emph{tensor spaces}
                \[ \hypercore \in \bigotimes_{\catenumeratorin} \rr^2 \, . \]
            \end{itemize}
        \end{definition}

        A propositional formula is also called a \emph{tensor of leg dimension $2$}.


    \end{frame}


    \section{Diagramatic Depiction}

    \begin{frame}{Curse of dimensionality}

        \begin{center}
            \it    In view of all that..., the many obstacles we appear to have surmounted, what casts the pall over our victory celebration?
            It is the curse of dimensionality, a malediction that has plagued the scientists from earliest days. - \text{Richard Bellman}
        \end{center}

        When having $\catorder$ variables (here: boolean variables like $\catvariableof{F}$ with $\catdim=2$), the number of states is
        \begin{align*}
            \prod_{\catenumerator\in\{0,...,\catorder\}} \catdimof{\catenumerator} =
            2^{\catorder} \, .
        \end{align*}
        When having $\catorder=1000$:
        \begin{align*}
            2^{1000} \approx 10^{301} \quad \Big(\text{Number of atoms in the universe: }\approx 10^{81}\Big)
        \end{align*}

        \begin{block}{Curse of dimensionality}
            The number of states increased exponentially with the number of variables.
        \end{block}

    \end{frame}

    \begin{frame}{Consequences}
        As consequence of the curse of dimensionality:
        \begin{itemize}
            \item {\bf Storage problem:} We cannot explicitly store a boolean to any state of the system.
            \item {\bf Visualization problem:} We cannot draw large tensors in the coordinatewise picture introduced so far.
        \end{itemize}

        The solution: Introduction of \emph{Tensor Networks}
        \begin{itemize}
            \item Diagramatic notation of tensors by wires (instead of coordinates)
            \item Contractions can be denoted by connections of the wires
        \end{itemize}

    \end{frame}

    \begin{frame}{Depiction of Tensors}
        We will depict
        \begin{itemize}
            \item \emph{tensors by blocks}
            \item affected \emph{categorical variables by open legs}
        \end{itemize}
        For example, given the atomic variables
        \[ \{\catvariableof{\atomenumerator} \, : \, \atomenumeratorin \} \]
        with respective dimensions $\catdimof{\atomenumerator}\in\mathbb{N}$ we depict a tensor
        \[ \hypercore \in \bigotimes_{\atomenumeratorin} \rr^{\catdimof{\atomenumerator}} \]
        as
        \begin{center}
            \input{./tikz-pics/tensor.tex}
        \end{center}
    \end{frame}


    \begin{frame}{Example: Depiction of Vectors}
        For example, a vector
        \[ V \in \rr^{m}\]
        with an index variable $\catvariable\in[m]$ is depicted by a block with a single open leg
        \begin{center}
            \input{./tikz-pics/vector.tex}
        \end{center}
    \end{frame}

%% To Syntax
%
%    \begin{frame}{Depiction of Tensors}
%        We will depict
%        \begin{itemize}
%            \item \emph{tensors by blocks}
%            \item affected \emph{categorical variables by open legs}
%        \end{itemize}
%        For example, given the atomic variables
%        \[ \{\catvariableof{\atomenumerator} \, : \, \atomenumeratorin \} \]
%        with respective dimensions $\catdimof{\atomenumerator}\in\mathbb{N}$ we depict a tensor
%        \[ \hypercore \in \bigotimes_{\atomenumeratorin} \rr^{\catdimof{\atomenumerator}} \]
%        as
%        \begin{center}
%            \input{./tikz_pics/tensor.tex}
%        \end{center}
%    \end{frame}
%
%
%    \begin{frame}{Example: Depiction of Vectors}
%        For example, a vector
%        \[ V \in \rr^{m}\]
%        with an index variable $\catvariable\in[m]$ is depicted by a block with a single open leg
%        \begin{center}
%            \input{./tikz_pics/vector.tex}
%        \end{center}
%    \end{frame}


\end{document}