\section{Inference algorithms}

Contractions of sub-networks are the local inference steps.
We review algorithms like AC-3 in that perspective and give with knowledge propagation a generic inference scheme.


%\sect{Deciding Entailment by Local Contractions}\label{subsec:LocalEntailment}

When having a Constraint Satisfaction Problem on a large number of variables, which are densely connected by constraint tensors, direct exploitation of the global entailment criterion in \theref{the:contCriterionLogEntailment} will be infeasible.
An alternative to deciding entailment by global operations is the use of local operations.
Here we interpret a part of the network (for example a single core) as an own knowledge base (with atomic formulas being the roots of the directed subgraph, that is potentially differing with the atoms in the global perspective) and perform entailment with respect to that.

\subsection{Manipulation of constraint tensor networks}

\todo{Reformulate inference algorithms based on the following operations:}
Leaving the contraction invariant:
\begin{itemize}
    \item[i] Adding the support of a sub-contraction of a graph $\secgraph\subset\graph$ and nodes $\secnodes\subset\nodes$, and drop some $\edge\in\secgraph$ with $\edge\subset\secnodes$.
    \item[ii] Partition an edge $\edge$ into two edges $\edgeof{0},\edgeof{1}$ decorated by $\hypercoreofat{\edgeof{0}}{\catvariableof{\edgeof{0}}},\hypercoreofat{\edgeof{1}}{\catvariableof{\edgeof{1}}}$, if
    \begin{align*}
        \hypercoreofat{\edge}{\catvariableof{\edge}}
        = \hypercoreofat{\edgeof{0}}{\catvariableof{\edgeof{0}}} \otimes \hypercoreofat{\edgeof{1}}{\catvariableof{\edgeof{1}}}
    \end{align*}
\end{itemize}
Monotonic decrease of solutions:
\begin{itemize}
    \item Adding constraint tensors.
    Thus, the network is satisfiable if an increased one is satisfiable.
    If a decreased one is unsatisfiable, then also the network is unsatisfiable.
\end{itemize}

Elementary networks are satisfiable when not containing the zero tensor.

\red{Need partial ordering for the monotonicity?}

\subsubsection{Examples}

\begin{example}[Modus ponens]
    When there is an edge
    \begin{align*}
        \edge
        = \innodes \cup \outnodes
    \end{align*}
    with
    \begin{align*}
        \hypercoreofat{\edge}{\catvariableof{\edge}}
        = \left(\bigwedge_{\node\in\innodes}\catvariableof{\node}\right) \Rightarrow \left(\bigwedge_{\node\in\outnodes}\catvariableof{\node}\right)
    \end{align*}
    and for each $\node\in\innodes$ an edge $\{\node\}$ with $\hypercoreofat{\{\node\}}{\catvariableof{\node}}=\tbasisat{\catvariableof{\node}}$, then:
    \begin{itemize}
        \item Choose $\secgraph=\left(\edge,\{\edge\}\cup\bigcup_{\node\in\innodes}\{\node\}\right)$, $\secnodes=\edge$
        \item Apply $i$ rule to replace $\secgraph$ with $\hypercoreofat{\edge}{\catvariableof{\edge}}=\bigtimes_{\node\in\edge}\tbasisat{\catvariableof{\node}}$
        \item Apply $ii$ rule to split into edges $\{\node\}$ for $\node\in\edge$ with $\tbasisat{\catvariableof{\node}}$
    \end{itemize}
\end{example}

\begin{example}[Resolution]
    When there are two clauses with a complementary literal can apply $i$ with the complementary literal variable left out of the open variables and get the resolvent clause.
    Notice that the old clauses cannot always be dropped, since they are no subsets of $\secnodes$-
\end{example}

%Show the grand resolution theorem based on the increase method.

\subsect{Monotonicity of Entailment}

\todo{Remove markov networks for monotonocity statements.
Understand as the proof of the rule $i$.
}

Vanishing local contractions provide sufficient but not necessary criterion to decide entailment, as we show in the next theorem.

\begin{theorem}[Monotonicity of Entailment]
    \label{the:monotonEntailment}
    For any Markov Network $\probtensor^{\graph}$ on the decorated hypergraph $\graph$ and any subgraph $\secgraph$, we have for any formula that $\probtensor^{\graph}\models\exformula$ if $\probtensor^{\secgraph}\models\exformula$.
\end{theorem}

%To show this theorem, we show the following lemma, that whether a contraction of non-negative tensors vanishes, a vanishing contraction of a subset of these tensors is a sufficient criterion.
To prove the theorem, we first establish the following lemma that states if a contraction of non-negative tensors vanishes, the vanishing of a contraction over a subset of these tensors is a sufficient criterion.

\begin{lemma}
    \label{lem:monotocityOfVanishingContractions}
    For any non-negative tensor network $\extnet$ on $\graph$ and $\secedges\subset\edges$ we have the following.
    For $\secgraph=(\secnodes,\secedges)$ with $\secnodes=\cup_{\edge\in\secedges}\edge$ and the tensor network $\tnetof{\secgraph}$ with tensors coinciding on $\secedges$ with those in $\extnet$ we have
    \begin{align*}
        \contraction{\extnet} = 0
    \end{align*}
    if $\contraction{\tnetof{\secgraph}}=0$.
\end{lemma}
\begin{proof}
    Since the tensor network $\tnetof{\secgraph}$ is non-negative, we have whenever $\contraction{\tnetof{\secgraph}}=0$ that
    \begin{align*}
        \contractionof{\tnetof{\secgraph}}{\catvariableof{\secnodes}} = \zerosat{\catvariableof{\secnodes}} \, .
    \end{align*}
    It follows with the commutation of contractions (see \theref{the:splittingContractions} in \charef{cha:messagePassing}), that
    \begin{align*}
        \contractionof{\extnet}{\catvariableof{\nodes}}
        &= \contractionof{
            \{\hypercoreof{\edge} \wcols \edge\in\edges/\secedges\}
            \cup \{\contractionof{\tnetof{\secgraph}}{\catvariableof{\secnodes}}\}
        }{\catvariableof{\nodes}} \\
        &=    \contractionof{
            \{\hypercoreof{\edge} \wcols \edge\in\edges/\secedges\}
            \cup \{\zerosat{\catvariableof{\secnodes}}\}
        }{\catvariableof{\nodes}} \\
        &= 0
    \end{align*}
    Thus, also the contraction of $\extnet$ vanishes in this case.
\end{proof}

\begin{proof}[Proof of \theref{the:monotonEntailment}]
    We use \lemref{lem:monotocityOfVanishingContractions} on the subset $\tnetof{\secgraph}$ of the cores $\extnet$ to the Markov Network $\probof{\graph}$, which itself defines the Markov Network $\probof{\secgraph}$.
    Whenever $\probof{\secgraph}\models\exformula$ for a formula $\exformula$, then we have by \theref{the:contCriterionLogEntailment}
    \begin{align*}
        \contraction{\tnetof{\secgraph} \cup\{\lnot\exformula\}} = 0 \, .
    \end{align*}
    It follows with \lemref{lem:monotocityOfVanishingContractions} that also
    \begin{align*}
        \contraction{\tnetof{\graph} \cup\{\lnot\exformula\}} = 0 \, .
    \end{align*}
    and therefore $\probof{\graph}\models\exformula$.
\end{proof}


\begin{remark}
    To make use of \theref{the:monotonEntailment} we can exploit any entailment criterion.
    However, there is no general statement about entailment possible, when the local entailment does not hold.
    \theref{the:monotonEntailment} therefore just provides a sufficient but not necessary criterion of entailment with respect to $\probtensor^{\graph}$.
\end{remark}



\subsection{Syntactical inference algorithms}

\todo{
    Clauses are tensors, which either shrink or trivialize when contracted with basis vectors.
}

\subsubsection{Forward Chaining in Horn Logics}

\todo{
    When all clauses definite and of length at most $2$, then both the $\onetuple{\nodes}$ and $\zerotuple{\nodes}$ satisfy the constraints.
    Forward chaining as copying length $1$ clauses to other clauses and contract them.
    When terminated all clauses are of lenght at most $2$, and definite if they initially where definite (then any clause of length $1$ is a fact).
}

\subsubsection{Resolution}

\begin{theorem}[Grand resolution theorem]
    When a constraint tensor network of clauses is unsatisfiable, then the empty clause is in the closure under resolution.
\end{theorem}
\begin{proof}
    We give a standard proof \cite{russell_artificial_2021} with the perspective of contractions.
    Having a clause tensor network we add basis vectors to get a elementary tensor network, which is satisfiable when we started with the closure under resolution.
\end{proof}

\subsection{Constraint Propagation}

\todo{
    Understand constraint propagation as an iterative application of the $(i)$ manipulation rule where messages are computed and contracted into memory cores.
    (Each step applies the rule twice: New message core computed and fused into old message core.)
    Call $\kcore$ memory core, which is updated by a message.
}

%\subsect{Message cores}

To store preliminary conclusions, we define auxiliary message cores storing constraints on variables $\edge\subset\nodes$.
They are understood as logical formulas to the atomization variables $\left(\catvariableof{\edge}=\catindexof{\edge}\right)$ of the respective formulas
\begin{align*}
    \kcoreofat{\edge}{\catvariableof{\edge}}
    = \bigvee_{\catindexof{\edge} \wcols \kcoreofat{\edge}{\indexedcatvariableof{\edge}}} \bigwedge_{\node\in\edge} \left(\catvariableof{\edge}=\catindexof{\edge}\right) \, .
\end{align*}

\begin{definition}\label{def:knowledgeCoreSoundComplete}
    Let $\extnet$ be a constraint satisfaction problem. % and $$ be a message core. % \left(\catvariableof{\edge}=\catindexof{\edge}\right)
    We say that a message core $\kcoreofat{\edge}{\catvariableof{\edge}}$ is sound for $\extnet$, if
    \begin{align*}
        \nonzerocirc\contractionof{\extnet}{\catvariableof{\edge}}  \prec \kcoreofat{\edge}{\catvariableof{\edge}}
    \end{align*}
    and complete for $\extnet$ if in addition
    \begin{align*}
        \nonzerocirc\contractionof{\extnet}{\catvariableof{\edge}}=\kcoreofat{\edge}{\catvariableof{\edge}} \, .
    \end{align*}
\end{definition}

We now provide a solution algorithm for constraint satisfaction problems by propagating local contractions.
The dynamic programming paradigm is implemented by the storage of partial entailment results in message cores.
We then iterate over local entailment checks, where we recursively add further entailment checks to be redone due to additional knowledge.
This local entailment scheme is called Knowledge Propagation and described in a generic way in \algoref{alg:knowledgePropagation}.

\red{
    We understand the support of local contractions as messages, which are tensors $\kcoreofat{\edge}{\catvariableof{\edge}}$ associated with message edges $\edge$.
    These message cores are propagated through the network, and summarize neighboring constraints.
}

\todo{
    Adjust constraint propagation to match tnreason message passing schemes:
    Queue contains
}

\begin{algorithm}[hbt!]
    \caption{Constraint Propagation}\label{alg:knowledgePropagation}
    \begin{algorithmic}
        \Require Constraint tensor network $\extnet$ on $\graph=(\nodes,\edges)$, message edges $\domainedges$ and a set $\arbset$ of subsets of $\edges$ for local propagation
        \Ensure Message cores $\kcoreofat{\edge}{\catvariableof{\edge}}$ for $\edge\in\domainedges$ with $\contractionof{\extnet}{\catvariableof{\edge}}\prec\kcoreofat{\edge}{\catvariableof{\edge}}$
        \iosepline
        \State
        \State Initialize for all $\edge\in\domainedges$:
        \begin{align*}
            \kcoreofat{\edge}{\catvariableof{\edge}}
            =\onesat{\catvariableof{\edge}}
        \end{align*}
        \State Initialize a queue
        \begin{align*}
            \graphqueue =
            \left\{
                (\edgesof{inference},\edgeof{m}) \wcols \edgesof{inference}\in\arbset \ncond \edgeof{m}\in\domainedges \ncond \edgeof{m} \cap \big(\bigcup_{\edge\in\edgesof{inference}}\edge\big) \neq \varnothing
            \right\}
        \end{align*}
        \While{$\graphqueue$ is not empty}
            \State Choose a set of edges from the queue
            \begin{align*}
                \inferenceedges,\edgeof{m} \algdefsymbol \graphqueue\mathrm{.pop()}
            \end{align*}
            %\ForAll{$\edge\in\domainedges$ with $\edge\cap\bigcup_{\secedge\in\secedges}\secedge\neq\varnothing$}
            \State Contract
            \begin{align*}
                \hypercoreat{\catvariableof{\messageedge}}
                = \nonzerocirc
                \contractionof{
                    \{\hypercoreofat{\secedge}{\catvariableof{\secedge}}\wcols\secedge\in\inferenceedges\}
                    \cup \{\kcoreofat{\edge}{\catvariableof{\edge}}\wcols\edge \in \domainedges , \, \edge\cap\bigcup_{\secedge\in\secedges}\secedge\neq\varnothing \}
                }{\catvariableof{\messageedge}}
            \end{align*}

%            \begin{align*}
%                \hypercoreat{\catvariableof{\edge}}
%                = \nonzerocirc\contractionof{
%                    \{\hypercoreofat{\secedge}{\catvariableof{\secedge}}\wcols\secedge\in\secedges\}
%                    \cup \{\kcoreofat{\edge}{\catvariableof{\edge}}\wcols\edge \in \domainedges , \, \edge\cap\bigcup_{\secedge\in\secedges}\secedge\neq\varnothing \}
%                }{\catvariableof{\edge}}
%            \end{align*}
            \If{$\hypercoreat{\catvariableof{\edge}}\neq\kcoreofat{\edge}{\catvariableof{\edge}}$}
                \begin{align*}
                    \kcoreofat{\edge}{\catvariableof{\edge}} \algdefsymbol \hypercoreat{\catvariableof{\edge}}
                \end{align*}
                \ForAll{$\secedges\in\arbset$ with $\edge\cap\bigcup_{\secedge\in\secedges}\secedge\neq\varnothing$}
                    \begin{align*}
                        \graphqueue\mathrm{.push(}\secedges\mathrm{)}
                    \end{align*}
                \EndFor
            \EndIf
        %\EndFor
        \EndWhile
        \State \Return $\{\kcoreofat{\edge}{\catvariableof{\edge}}\,:\,\edge\in\domainedges\}$
    \end{algorithmic}
\end{algorithm}

% Interpretation
Each chosen subset $\secedges\in\arbset$ is understood as a local knowledge base, which is then applied for local entailment.
The message cores are understood as messages, which propagate information from different regions of a tensor network (see \charef{cha:messagePassing}).

%Implementation
There are different ways of implementing \algoref{alg:knowledgePropagation}, by choosing the set $\arbset$ of constraint sets $\secedges$ and domain $\domainedges$. % and a stopping criterion.


\input{partI/0-propositional-logics/examples/inference-algorithms/ac-3}


%The central property used in knowledge propagation is that any subcontraction can be added to the constraint network without changing it.
%\begin{theorem}\label{the:booleanContractionInvariance}
%	Given a boolean tensor network $\extnet$ on $\graph$ and $\secedges\subset\edges$.
%	For $\secgraph=(\secnodes,\secedges)$ with $\nodes=\cup_{\edge\in\secedges}\edge$ and the tensor network $\tnetof{\secgraph}$ with boolean tensors coinciding on $\secedges$ with those in $\extnet$ we have
%	\begin{align*}
%		\contractionof{\extnet}{\catvariableof{\nodes}} =
%		\contractionof{\extnet\cup\{\nonzerocirc\contractionof{\tnetof{\secgraph}}{\thirdnodes}\}}{\catvariableof{\nodes}} \, ,
%	\end{align*}
%	where $\thirdnodes\subset\nodes$ is arbitrary.
%\end{theorem}
%\begin{proof}
%	We will proof this statement later in \charef{cha:messagePassing}, see \theref{the:invarianceAddingSubcontractions}.
%\end{proof}

\begin{theorem}
    \label{the:soundnessKnowledgePropagation}
    At any state of the Knowledge Propagation \algoref{alg:knowledgePropagation}, we have that each message core $\kcoreof{\secedge}$ is sound for $\extnet$.
    After each update in \algoref{alg:knowledgePropagation}, $\kcoreof{\secedge}$ is further monotonically decreasing with respect to the partial ordering.
\end{theorem}
\begin{proof}
    \todo{This follows from soundness of the $i$ manipulation rule.}
    We show the first claim by induction over the update steps in \algoref{alg:knowledgePropagation}.
    At the start, where $\kcoreofat{\edge}{\edgevariables} = \onesat{\edgevariables}$, we trivially have
    \begin{align*}
        \contractionof{\extnet \cup \{\kcoreofat{\edge}{\edgevariables} \wcols \edge\in\domainedges\}}{\nodevariables}
        = \contractionof{\extnet \cup \{\onesat{\edgevariables} \wcols \edge\in\domainedges\}}{\nodevariables}
        = \contractionof{\extnet}{\nodevariables} \, .
    \end{align*}
    Let us now assume, that for a state of cores $\{\kcoreof{\edge} \wcols \edge\in\domainedges\}$ the first claim holds and let $\secedges\subset\edges$ be chosen for the update of $\kcoreof{\secedge}$.
    By the invariance under adding the support of subcontractions, which we will proof in more detail as \theref{the:monotonicityBinaryContractions} in \charef{cha:messagePassing}, we have for the update
    \begin{align*}
        \tilde{\kcore}^{\secedge}[\catvariableof{\secedge}]
        = \nonzerocirc\contractionof{
            \{\hypercoreof{\edge} \wcols \edge\in\secedges\} \cup \{\kcoreof{\edge} \wcols \edge\in\domainedges, \, \edge\cap\bigcup_{\secedge\in\secedges}\secedge\neq\varnothing  \}
        }{\catvariableof{\secedge}}
    \end{align*}
    that
    \begin{align*}
        \contractionof{\extnet \cup \{\kcoreofat{\edge}{\edgevariables} \wcols \edge\in\domainedges\}}{\nodevariables}
        = \contractionof{\extnet \cup \{\kcoreofat{\edge}{\edgevariables} \wcols \edge\in\domainedges\} \cup \{\tilde{\kcore}^{\secedge}[\catvariableof{\secedge}]\}
        }{\nodevariables}  \, .
    \end{align*}
    Thus, the first claim holds also after the update of the core to $\secedge$.

    % Since the kcore is in the contraction
    We further have with the monotonicity of boolean contraction (see \theref{the:monotonicityBinaryContractions}) that for any update of $\kcoreof{\secedge}$ by $\tilde{\kcore}^{\secedge}$
    \begin{align*}
        \tilde{\kcore}^{\secedge}[\catvariableof{\secedge}]
        = \nonzerocirc\contractionof{
            \{\hypercoreof{\edge} \wcols \edge\in\secedges\}
            \cup \{\kcoreof{\edge} \wcols \edge\in\domainedges, \, \edge\cap\bigcup_{\edge\in\secedges}\edge\neq\varnothing\}
        }{\catvariableof{\secedge}}
        \prec \kcoreofat{\secedge}{\catvariableof{\secedge}} \, .
    \end{align*}
    Thus, each message core is monotonously decreasing at each update, with respect to the partial tensor ordering.

    From the first claim we further have for any $\secedge\in\secedges$
    \begin{align*}
        \contractionof{\{\kcoreofat{\secedge}{\catvariableof{\secedge}}\}\cup\extnet \cup \left\{\kcoreof{\edge} \wcols \edge\in\domainedges/\{\secedge\}\right\}}{\catvariableof{\secedge}}
        = \contractionof{\extnet}{\catvariableof{\secedge}}
    \end{align*}
    And thus in combination with the monotonicity of boolean contraction (see \theref{the:monotonicityBinaryContractions}) that
    \begin{align*}
        \nonzeroof{\contractionof{\extnet}{\catvariableof{\secedge}}} \prec \kcoreofat{\secedge}{\catvariableof{\secedge}}  \, . & \qedhere
    \end{align*}
\end{proof}

Let us now show that Knowledge Propagation always terminates.
We can further characterize the message cores at termination.

\begin{definition}
    We say that a set of message cores $\{\kcoreof{\edge} \wcols \edge\in\domainedges\}$ is consistent with a set $\{\hypercoreof{\edge} \wcols \edge\in\secedges\}$, if for any $\edge\in\domainedges$
    \begin{align*}
        \kcoreofat{\edge}{\catvariableof{\edge}}
        = \nzcontractionof{\{\hypercoreof{\edge} \wcols \edge\in\secedges\} \cup \{\kcoreof{\edge} \wcols \edge\in\domainedges\}}{\catvariableof{\edge}} \, .
    \end{align*}
\end{definition}

This property is similar to the completeness of a message core, when interpreting the other message cores and the constraints $\{\hypercoreof{\edge} \wcols \edge\in\secedges\}$ as posing a Constraint Satisfaction Problem.

\begin{theorem}
    Knowledge Propagation \algoref{alg:knowledgePropagation} always terminates.
%    At termination, each message core $\kcoreof{}$
    At termination we further for each $\secedges\in\arbset$ and $\edge\in\domainedges$ with $\edge\cap\bigcup_{\edge\in\secedges}\edge\neq\varnothing$, that the message cores $\{\kcoreof{\edge} \wcols \edge\in\domainedges, \, \edge\cap\bigcup_{\edge\in\secedges}\edge\neq\varnothing\}$ are consistent with $\{\hypercoreof{\edge} \wcols \edge\in\secedges\}$.
\end{theorem}
\begin{proof}
    For each message core, there are finitely many boolean tensor precessing it with respect to the partial order.
    Therefore, since they are monotonously decreasing, each message core can only be varied finitely many times during the algorithm.
    In total the algorithm can run only finitely many times in the second for loop, where new sets of edges are pushed into the queue.
    Therefore the while loop will always terminate.

    When after a single pass through the while loop with chosen $\secedges\in\arbset$, the set $\secedges$ is not pushed back into $\graphqueue$, we have for any $\edge\in\domainedges$ with $\edge\cap\bigcup_{\edge\in\secedges}\edge\neq\varnothing$ that
%    At termination, we have for any $\secedges\in\arbset$
    \begin{align*}
        \kcoreofat{\edge}{\catvariableof{\edge}}
        = \nzcontractionof{\{\hypercoreof{\edge} \wcols \edge\in\secedges\} \cup \{\kcoreof{\edge} \wcols \edge\in\domainedges, \, \edge\cap\bigcup_{\edge\in\secedges}\edge\neq\varnothing\}}{\catvariableof{\edge}} \, .
    \end{align*}
    Whenever the contraction on the right hand side changes during the algorithm, the set $\secedges$ is pushed into $\graphqueue$.
    At termination of the algorithm, $\graphqueue$ is empty, and the claimed consistency therefore has to hold.
%    since otherwise the queue $\graphqueue$ would at least contain $\secedges$ and the while loop would not terminate.
\end{proof}

%A direct consequence of this theorem is that when running Knowledge Propagation with all edges, then
%\begin{corollary}
%    When we choose $\arbset=\{\edges\}$ for an CSP $\extnet$, then the message cores returned by \algoref{alg:knowledgePropagation} are complete.
%\end{corollary}


We can exploit the Knowledge Propagation \algoref{alg:knowledgePropagation} for the solution of Constraint Satisfaction Problems, by taking $\extnet$ as the tensor network of constraint tensors.
Whenever a message core vanishes, we can conclude that the Constraint Satisfaction Problems is not satisfiable, as we show next.

\begin{corollary}
    Let us for a Constraint Satisfaction Problem encoded by $\extnet$ run Knowledge Propagation \algoref{alg:knowledgePropagation}.
    Whenever for any $\secedge\in\edges$ we have $\kcoreofat{\secedge}{\catvariableof{\secedge}}=\zerosat{\catvariableof{\secedge}}$, then the Constraint Satisfaction Problem is not satisfiable.
\end{corollary}
\begin{proof}
    Whenever $\kcoreofat{\secedge}{\catvariableof{\secedge}}=\zerosat{\catvariableof{\secedge}}$, then we have by \theref{the:soundnessKnowledgePropagation}
    \begin{align*}
        \nonzeroof{\contractionof{\extnet}{\catvariableof{\secedge}}} \prec \zerosat{\catvariableof{\secedge}}
    \end{align*}
    and therefore
    \begin{align*}
        \contraction{\extnet} = 0 \, . & \qedhere
    \end{align*}
\end{proof}

% Combination with backtracking search
When the Knowledge Propagation \algoref{alg:knowledgePropagation} converges in a given implementation and no message core vanishes, we can however not conclude that the Constraint Satisfaction Problem is not satisfiable.
However, for any index tuple $\catindexof{\nodes}$ to be a solution of the CSP to $\extnet$, we have the necessary condition
\begin{align*}
    \uniquantwrtof{\secedge\in\secedges}{\kcoreofat{\edge}{\catvariableof{\secedge}=\restrictionofto{\catindexof{\nodes}}{\secedges}}=1} \, ,
\end{align*}
where by $\restrictionofto{\catindexof{\nodes}}{\secedges}$ we denote the restriction of the index tuple $\catindexof{\nodes}$ to the variables included in $\secedges$.
One can use this insight as a starting point for backtracking search, where the assignments to variables $\catvariableof{\secnodes}$ are iteratively guessed, based on the restriction that each constraint is locally satisfialbe, i.e. .
\begin{align*}
    \uniquantwrtof{\secedge\in\secedges}{
        \contraction{\kcoreofat{\edge}{
            \catvariableof{\secedge\cap\secnodes}=\restrictionofto{\catindexof{\nodes}}{\secedge\cap\secnodes}
            ,\catvariableof{\secedge/\secnodes}}} \neq 0
    } \, .
\end{align*}
One can understand the guess of an assignment $\catindexof{\node}$ to a variable $\catvariableof{\node}$, as it is done during backtracking search, as an inclusion of a constraint
\begin{align*}
    \kcoreofat{\{\node\}}{\catvariableof{\node}}
    = \onehotmapofat{\catindexof{\node}}{\catvariableof{\node}} \, .
\end{align*}
Therefore, Knowledge Propagation \algoref{alg:knowledgePropagation} can be integrated with backtracking search, with iterations between propagations of knowledge and guessing of additional variables.
%The message cores $\kcoreof{\edge}$ are subset encoding of possible local choices, according to which variables can be assigned.


\subsect{Applications}

Let us exemplify the usage of Knowledge Propagation on Constraint Satisfaction Problems posed by entailment queries on Markov Networks.

\begin{corollary}
    \label{cor:knowledgePropagationMarkovNetworks}
    Let \algoref{alg:knowledgePropagation} be run on the cores $\tnetof{\graph}\cup\{\bencodingof{\exformula}\}$ with an arbitrary design of $\secedges$.
    Whenever for a formula $\formulaat{\catvariableof{\secnodes}}$ and a $\kcoreof{\edge}$ we have
    \[ \contractionof{\kcoreof{\edge},\bencodingof{\exformula}}{\exformulavar=0} =0  \]
    then the Markov Network $\extnet$ probabilistically entails $\exformula$.
    If on the contrary
    \[ \contractionof{\kcoreof{\edge},\bencodingof{\exformula}}{\exformulavar=1} =0  \]
    then the Markov Network $\extnet$ probabilistically entails $\lnot\exformula$, that is probabilistically contradicts $\exformula$.
\end{corollary}
\begin{proof}
    This follows from \theref{the:soundnessKnowledgePropagation} ensuring the soundness of Knowledge Propagation and the sufficiency of local entailment.
\end{proof}

\begin{example}[Batch decision of entailment]
    Let $\formulaset$ be a set of formulas and $\probofat{\graph}{\shortcatvariables}$ a Markov Network, for which it shall be decided, which formulas in $\formulaset$ are entailed, contradicted or contingent.
    We can in addition to the cores of the Markov Network create the cores $\{\bencodingofat{\exformula}{\formulavar,\shortcatvariables} \wcols \exformula\in\formulaset\}$ and prepare the message cores
    \begin{align*}
        \kcoreofat{\{\formula\}}{\formulavar} \, .
    \end{align*}
    To decide entailment batchwise, Knowledge Propagation \algoref{alg:knowledgePropagation} can be run.
    Whenever during the algorithm we have that for a $\formula$, then \corref{cor:knowledgePropagationMarkovNetworks} implies that if
    \begin{align*}
        \kcoreofat{\{\formula\}}{\formulavar} =
        \begin{cases}
            \tbasisat{\formulavar} & \text{then, the formula is entailed by }\probof{\graph} \, . \\
            \fbasisat{\formulavar} & \text{then, the formula is contradicted by }\probof{\graph} \, .\\
            \onesat{\formulavar} & \text{then no conclusion can be drawn.}
        \end{cases}
    \end{align*}
    Note, that $\kcoreofat{\{\formula\}}{\formulavar} = \zerosat{\formulavar}$ can not happen, since this would mean that $\nonzeroof{\probof{\graph}}$ is inconsistent.
    Thus, at any stage of \algoref{alg:knowledgePropagation}, one of the three holds.
    % Inference rules such as modus ponens can be mimicked.
\end{example}


\subsect{Mimicking Inference Rules by Propagation}

While so far we have discussed semantic based entailment, there are inference rules exploiting only logical syntax to infer entailed statements.
We here show, that they can be captures by the knowledge propagation scheme, if the sets $\arbset$ and $\domainedges$ are chosen properly.

Whenever
\begin{align*}
    \bigvee_{\exformula\in\formulaset} \exformula \models \secexformula
\end{align*}
then
\begin{align*}
    \tbasisat{\secexformulavar} =
    \contractionof{\{\formulaat{\shortcatvariables} \wcols \formula\in\formulaset\} \cup \{\bencodingofat{\secexformula}{\secexformulavar,\shortcatvariables}\}}{\secexformulavar} \, ,
\end{align*}
that is the inference rule can be performed when $\formulaset$ are in $\arbset$.

\begin{example}{Modus Ponens}
    For example, when for two formulas $\exformula,\secexformula\in\formulaset$ we have $\exformula\models\secexformula$, then when $\kcoreofat{\{\exformula\}}{\exformulavar} = \tbasisat{\formulavar}$ we have
    \begin{align*}
        \tbasisat{\secexformulavar}
        = \contractionof{(\exformula\Rightarrow\secexformula)[\exformulavar,\secexformulavar],\kcoreofat{\{\exformula\}}{\exformulavar}}{\secexformulavar} \, ,
    \end{align*}
    that is entailment of $\secexformula$ can be concluded using a single update.

    When we have a Knowledge Base of horn clauses, we run Knowledge Propagation with each horn clause being a constraint core and a message core for any variable.
    \algoref{alg:knowledgePropagation} therefore resembles the forward chaining algorithm of propositional logics (see Figure~7.15 in \cite{russell_artificial_2021}).
    It is known, that forward chaining is complete for Horn Logic.
    Thus, the message cores returned in that case by \algoref{alg:knowledgePropagation} are complete for the Knowledge Base as a CSP.
\end{example}

\subsection{Discussion}

In our perspective, the number of possible sub-networks to be contracted is the bottleneck, instead of the demand to perform such contractions.
%The number of sub-networks grows as binomial coefficients.
The choice and the scheduling of these sub-contractions is most important for the design of efficient inference schemes.
