\documentclass{scrartcl}
\input{../../../spec_files/exercise_preamble}

\begin{document}
    \exercisehead{1}{Modeling with Propositional Logic}{March 23, 2026}
    \usecasebox{Alarm System}{
        Imagine you want to estimate whether a Burglary or an Earthquake happened at your home, depending on whether your Alarm rings and whether your neighbors John and Mary called you.
        You know that:
        \begin{itemize}
            \item When a Burglary or an Earthquake occurs, the Alarm rings.
            \item When the Alarm rings then we receive a call from at least one of the neighbors John and Mary.
            \item The neighbor John never calls.
        \end{itemize}
    }
    \addtask{1 (after lecture I.2-Entailment)}{
    Define suitable Boolean variables for a factored representation of this system and express the knowledge by propositional formulas.
    Draw the tensor network diagram representing the knowledge base.
    }
    \addtask{2 (after lecture I.2-Entailment)}{
    Find the truth table representation and the tensor representation of each formula in the knowledge base.
    }
    \addtask{3 (after lecture I.2-Entailment)}{
    Decide which of these statements are entailed by our knowledge:
    \begin{itemize}
        \item When the Alarm rings, either a Burglary or an Earthquake happened.
        \item When the Alarm rings, then Mary calls.
        \item When neither Mary nor John calls, then either the alarm did not ring, or a Burglary did not happen, or an Earthquake did not happen.
    \end{itemize}
    }
    \addtask{4 (after lecture I.3-Syntax)}{
    Consider the syntactical decomposition of the formulas in the knowledge base into circuits of logical connectives.
    Draw the tensor network decomposing the basis encodings of the formulas.
    Find the coordinatewise description of the basis encodings to each appearing connective.
    }
\end{document}