\documentclass{beamer}

\input{../../spec_files/course_preamble.tex}
\input{../../spec_files/course_macros.tex}

\title[Circuits]{
    \coursechapterone \\
    {\huge Circuits}
}

\begin{document}

    {\frame[plain]{\titlepage}}

    \showoutlineatsection


    \section{Motivation}

    \begin{frame}{Propositional Syntax}

        Propositional syntax decomposes formulas into connectives applied on sub-formulas.

        When understanding formulas as functions, we get a decomposition into circuits of connectives.

        This decomposition is reflected in the tensor network decomposition of basis encodings.

        \begin{center}
            \bf Can we generalize to arbitrary functions and find tensor network decompositions based on circuits?
        \end{center}

        Example: Neural networks are circuits of neurons.

    \end{frame}

    \begin{frame}{Circuits as a computational model}

        Circuits are a fundamental computational model, which is used in many areas of computer science, e.g., in complexity theory, logic, and machine learning.

    \end{frame}


    \section{Example: $\catdim$-adic sums}

    \begin{frame}{$\catdim$-adic representation of integers}

        Tensor: Each coordinate is the integer to the $\catdim$-adic representation given by the indices.

        How to decompose the sum tensors, which are tensors with legs to both variables?

    \end{frame}

    \begin{frame}{Decomposition into half-adders}

        Summation can be done by concatenation of half-adders, which are circuits with two input legs and two output legs, one for the sum and one for the carry.

    \end{frame}

    \begin{frame}{Extension to multiplications}
        Do the multiplication for each digit of the right number separately, then add the results.

        In this way: Generalize to

    \end{frame}


    \section{General circuits}

    \begin{frame}{Set enumerating variables}

        Generic basis encodings of functions contain set enumerating variables to the input and target sets.

    \end{frame}

    \begin{frame}{Circuit}

        A circuit is a decorated hypergraph.

    \end{frame}

    \begin{frame}{Decomposition into tensor networks}

        The basis encoding to a function of a circuit is decomposed into basis encodings of the node functions.

    \end{frame}


\end{document}













